---
title: "Using Bayesian radiocarbon chronology package Bchron"
author: "Andrew Parnell"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bchron}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

Bchron is an R package that enables quick calibration of radiocarbon dates under various calibration curves (including user generated ones); age-depth modelling as per the algorithm of Haslett and Parnell (2008); Relative sea level rate estimation incorporating time uncertainty in polynomial regression models (Parnell and Gehrels 2015); and non-parametric phase modelling via Gaussian mixtures as a means to determine the activity of a site (and as an alternative to the Oxcal function SUM; currently unpublished).

You will find Bchron far easier to use if you know some basics on how to use R. I recommend the book by Norman Matloff [amazon.co.uk link](http://www.amazon.co.uk/The-Art-Programming-Statistical-Software/dp/1593273843), or the online intro course by [Code School](www.codeschool.com).

If you find bugs or want to suggest new features please visit the Bchron [GitHub issues page](https://github.com/andrewcparnell/Bchron/issues). 

## Installing Bchron

Bchron will run in Windows, Mac OS X or Linux. To install Bchron you first need to install [R](http://www.r-project.org). I would also recommend installing [Rstudio](http://www.rstudio.com) as a nice desktop environment for using R. Once in R you can type:

```{r,eval=FALSE}
install.packages('Bchron')
```

at the R command prompt to install Bchron. If you then type:

```{r}
library(Bchron)
```

it will load in all the Bchron functions.

## Calibrating radiocarbon dates

Bchron will calibrate single or multiple dates under multiple (even user defined) calibration curves. By default, the intcal13, shcal13 and marine13 calibration curves are included. You can calibrate a single radiocarbon date with, e.g.

```{r,fig.align='center',fig.width=6,fig.height=5}
ages1 = BchronCalibrate(ages=11553,
                        ageSds=230,
                        calCurves='intcal13',
                        ids='Date-1')
summary(ages1)
plot(ages1)
```

This will calibrate the radiocarbon age of 11,553 14C years BP with standard error 230 14C years on the intcal13 calibration curve. The `id` given is optional and only used for summarising and plotting. The summary command then gives the highest density regions of the calibrated date and the plot command produces a simple plot of the density.

Bchron can calibrate multiple dates simultaneously by inputting the dates as vectors:

```{r,results='hide'}
ages2 = BchronCalibrate(ages=c(3445,11553,7456), 
                        ageSds=c(50,230,110), 
                        calCurves=c('intcal13','intcal13','shcal13'))
summary(ages2)
plot(ages2)
```

This will calibrate three different 14C ages with the calibration curves as specified in the `calCurves` argument. The `summary` and `plot` commands will produce individual highest density regions and density plots for the three dates.

Finally, if you provide position information (e.g. depths) to the BchronCalibrate function it will create a plot with position on the y-axis, e.g.:

```{r,fig.align='center',fig.width=6,fig.height=5}
ages3 = BchronCalibrate(ages=c(3445,11553), 
                        ageSds=c(50,230), 
                        positions=c(100,150), 
                        calCurves=c('intcal13','normal'))
summary(ages3)
plot(ages3,withPositions=TRUE)
```

If, alternatively, you want a _credible interval_ (i.e. a single contiguous range) rather than an HDR, you could use the `SampleAges` function:

```{r}
# First create age samples for each date
age_samples = SampleAges(ages3)
# Now summarise them with quantile - this gives a 95% credible interval
apply(age_samples,2,quantile,prob=c(0.025,0.975))
```

The calibration code is very fast. On standard PC you should have no trouble calibrating thousands of dates simultaneously without a noticeably long wait.

## Running the Bchronology age-depth model

The Bchronology function fits the age-depth model outlined by Haslett and Parnell (2008). An illustrative data set from Glendalough is provided with the package, containing 5 radiocarbon dates and a known age for the top of the core. It can be called in via:

```{r}
data(Glendalough)
print(Glendalough)
```

The top date is from the present and has the calibration curve 'normal' as it is not a 14C date. This core can be run through Bchron via:

```{r,results='hide'}
GlenOut = Bchronology(ages=Glendalough$ages,
                      ageSds=Glendalough$ageSds, 
                      calCurves=Glendalough$calCurves,
                      positions=Glendalough$position, 
                      positionThicknesses=Glendalough$thickness,
                      ids=Glendalough$id, 
                      predictPositions=seq(0,1500,by=10))
```

There are other arguments you can supply to Bchronology, including the date the core was extracted, the outlier probabilities for each individual date, and the number of iterations for which to run the algorithm. For more details see:

```{r,eval=FALSE}
help(Bchronology)
```

Once run, the summary commands will show various output:
```{r,eval=FALSE}
summary(GlenOut)
```
```{r}
summary(GlenOut, type='convergence')
summary(GlenOut, type='outliers')
```

The first summary command produces ages for each position supplied in the `predictPositions` argument above (output not shown as it's too long). The second provides convergence diagnostics. The third gives outlier probabilities. The `plot` command will produce an age-position plot:

```{r,fig.align='center',fig.width=6,fig.height=5}
plot(GlenOut,
     main="Glendalough",
     xlab='Age (cal years BP)',
     ylab='Depth (cm)',
     las=1)
```

Finally, the `predict` command will produce predicted ages for a newly specified set of positions with optional thicknesses:

```{r,results='hide'}
predictAges = predict(GlenOut, 
                      newPositions = c(150,725,1500), 
                      newPositionThicknesses=c(5,0,20))
```

To run this model on a data set of your own, you will need to load in your data set via, e.g.

```
mydata = read.table(file='path/to/file.txt',header=TRUE)
run = Bchronology(ages=mydata[,1],ageSds=mydata[,2], ...
```

## Running RSL rate estimation

The function `BchronRSL` will produce estimated relative sea level rates from a regression model taking into account the uncertainties in age provided by a `Bchronology` run as above. Two example data sets are provided:

```{r}
data(TestChronData)
data(TestRSLData)
```


These can be run through Bchronology and BchronRSL via:

```{r,eval=FALSE}
RSLrun = Bchronology(ages=TestChronData$ages,
                     ageSds=TestChronData$ageSds, 
                     positions=TestChronData$position,
                     positionThicknesses=TestChronData$thickness, 
                     ids=TestChronData$id,
                     calCurves=TestChronData$calCurves, 
                     predictPositions=TestRSLData$Depth)
RSLrun2 = BchronRSL(RSLrun,
                    RSLmean=TestRSLData$RSL,
                    RSLsd=TestRSLData$Sigma,
                    degree=3)
```

The `Bchronology` run is as described in the section above. The `BChronRSL` run takes this object, an estimate of the RSL means and standard deviations, and a value of degree (here 3 indicating cubic regression). They can then be summarised and plotted via:

```{r,fig.align='center',fig.width=6,fig.height=5,eval=FALSE}
summary(RSLrun2)
plot(RSLrun2)
```

## Running non-parametric phase estimation

Bchron contains two functions for running non-parametric phase models for estimating activity level in a site/region. The first, BchronDensity fits a full Bayesian Gaussian mixture model to the radiocarbon dates whilst the second BchronDensityFast fits an approximate version which will run on much larger data sets. An example run is

```{r,results='hide'}
data(Sluggan)
SlugDens = BchronDensity(ages=Sluggan$ages,
                         ageSds=Sluggan$ageSds,
                         calCurves=Sluggan$calCurves,
                         numMix=50)
```
The `BchronDensity` function also has an optional argument `store_density` which will output an age grid and associated density in case you wish to plot the output in a different package.


```{r,fig.align='center',fig.width=6,fig.height=5}
plot(SlugDens,xlab='Age (cal years BP)')
```

BchronDensityFast is identical except for the function call:

```{r,eval=FALSE}
SlugDensFast = BchronDensityFast(ages=Sluggan$ages,
                                 ageSds=Sluggan$ageSds, 
                                 calCurves=Sluggan$calCurves)
```
The `BchronDensityFast` function outputs the age grid and density by default so requires no extra arguments to store it.

## Including user-defined calibration curves

Bchron allows for user-defined calibration curves to be created through the function `CreateCalCurve`. If inputs to this function are:
  - The `name` of the calibration curve (e.g. intcal13c)
  - The calibrated/calendar ages for the new calibration curve
  - The uncalibrated (e.g. 14C) ages for the new calibration curve
  - (Optionally) the 1-sigma error for the calibration curve in uncalibrated years. If not included the error is assumed to be zero

One run, the calibration curve can be used for all future runs until the package is deleted or re-installed. The calibration curve can be used by simple calling it in any other Bchron function using the supplied name. Note that no name checking is done, so it is possible to over-write the pre-defined calibration curves if

As an example, consider installing the intcal09 calibration curve, as available from (http://www.radiocarbon.org/IntCal09%20files/intcal09.14c)[here]. We can run the following code:

```{r}
# Load in the calibration curve with:
intcal09 = read.table('http://www.radiocarbon.org/IntCal09%20files/intcal09.14c',sep=',')
# Run CreateCalCurve
CreateCalCurve(name='intcal09',cal_ages=intcal09[,1],uncal_ages=intcal09[,2],one_sigma=intcal09[,3])
```

The new calibration curve can now be used via its name. For example, we can now compare the calibration of a date under the intcal09 curve with that of intcal13:
```{r}
age_09 = BchronCalibrate(age=15500,ageSds=150,calCurves = 'intcal09',ids='My Date')
age_13 = BchronCalibrate(age=15500,ageSds=150,calCurves = 'intcal13')
plot(age_09)
with(age_13$date1,lines(ageGrid,densities,col='red'))
legend('topleft',legend=c('intcal09','intcal13'),col=c('black','red'),lty=1)
```

## References

For a description of the kind of things that Bchron can do:  
Parnell, A. C., Haslett, J., Allen, J. R. M., Buck, C. E., & Huntley, B. (2008). A flexible approach to assessing synchroneity of past events using Bayesian reconstructions of sedimentation history. Quaternary Science Reviews, 27(19-20), 1872–1885.

For the maths behind Bchron:  
Haslett, J., & Parnell, A. (2008). A simple monotone process with application to radiocarbon-dated depth chronologies. Journal of the Royal Statistical Society: Series C (Applied Statistics), 57(4), 399–418.

For a review of chronology models:  
Parnell, A. C., Buck, C. E., & Doan, T. K. (2011). A review of statistical chronology models for high-resolution, proxy-based Holocene palaeoenvironmental reconstruction. Quaternary Science Reviews, 30(21-22), 2948–2960.

For relative sea level rate estimation:  
Parnell, A. C., & Gehrels, W. R. (2015). Using chronological models in late Holocene sea level reconstructions from salt marsh sediments. In: I. Shennan, B.P. Horton, and A.J. Long (eds). Handbook of Sea Level Research. Chichester: Wiley.

